Rising from the Ashes II: Did the Birth of the Milky Way's Bar Give Birth to the Birth of the Abundance Bimodality

### 2024-09-10 paper/ms.tex

\begin{abstract}
    The Milky Way is known to host at least two modes in the present day distribution of Fe and $\alpha$-elements. The exact cause of this bimodality is disputed, but one class of explanations involves the merger between the Milky Way and a relatively massive satellite (Gaia-Sausage-Enceladus) at $z\sim2$. However, reproducing this bimodality in simulations is not straightforward, with conflicting results on the prevalance, morphology, and mechanism behind multimodality. We present a case study of a galaxy in the Illustris TNG50 simulation which undergoes a sequence of starburst, brief quiescence, and then rejuvenation. After a minor post-processing step which boosts the \alphaFe{} of old star particles, we demonstrate that this galaxy hosts a strongly bimodal distribution in the \alphaFe{}-\FeH{} plane. The high- and low-$\alpha$ sequences are neatly separated in time by the brief quiescent period. The quiescent period in this galaxy is not associated with a merger but by the formation of a bar which is followed by AGN activity. This galaxy indicates a novel scenario in which the $\alpha$-bimodality is caused by AGN-induced quenching induced by the formation of the Milky Way's bar. We argue that the post-processing step can be interpreted as the TNG model underproducing star formation in the densest regions at high-$z$.
  \end{abstract}

@@

In that work, we used idealized simulations of a galaxy merger that triggered a starburst which preceded the quiescent period. However, we argued that the merger aspect of that work was not necessary, but rather the quiescence. In this work, we study a subhalo from the Illustris TNG50 simulation which demonstrates this. This subhalo of interest (SoI) exhibits the sequence of events presented in \citet{2024arXiv240707985B} after a simple post-processing step which enhances the \alphaFe{} of old star particles. The SoI undergoes a brief quiescent period which neatly separates a high- and low-$\alpha$ sequence. Therefore, this work serves as a verification that the scenario in \citet{2024arXiv240707985B} is possible in cosmological simulations and can occur in non-merger scenarios.

In Section~\ref{sec:methods} we discuss our selection technique which led to discovering the SoI, as well as a simple one zone chemical evolution model we use to justify our post-processing step. In Section~\ref{sec:results} we present the main results which we discuss and interpret in Section~\ref{sec:disc}. We conclude in Section~\ref{sec:conc}.

@@

Using the public catalog, we selected a sample of subhalos at $z=1.5$ (snapshot 40) according to the following criteria: (1) the subhalo is central (i.e., the most massive subhalo within its halo), and (2) the subhalo's stellar mass is between $10^{10}$ and $10^{10.5}\,\Msun/\textrm{h}$. There were a total of 168 subhalos that met both criteria. The chosen mass range is broadly consistent with the expected mass of the Milky Way at this redshift \citep{2013ApJ...771L..35V}. We chose to make our selection of galaxies at $z=1.5$ instead of at lower redshift because we wished to capture the \textit{formation} of any multimodal structure. We did not want contamination by mergers at lower redshift which we know contribute very little to the Milky Way's disk stars \citep[e.g.,][]{2016ARA&A..54..529B}. We study the main descendant of our SoI at $z=0$

@@

@@

The colored vertical lines are indicated at $\FeH=-0.75$, $-0.5$, $-0.25$, and $0\,\dex$ in the Milky Way, and at bins $0.25\,\dex$ higher in the simulations. The lower panels show 1D histograms of \MgFe{} in bins centered on these values. The bins for the Milky Way/simulations are $0.2$/$0.05\,\dex$. The \MgFe{} values are given offsets so their medians are at zero. The rationale for the higher plotted \FeH{} in the simulations comes just from the empirical location of the bimodalities. The Milky Way shows a clear bimodal population, with a high-$\alpha$ sequence most clearly distinct from the low-$\alpha$ sequence at low metallicity. The two sequences merge around solar metallicity.

@@

In the lower panels we plot to the left data from the Milky Way and to the middle and right from the fiducial and $\alpha$-enhanced SoI, respectively. For the simulations, we add $15\%$ age and $0.01\,\dex$ \MgFe{} Gaussian errors. These values are roughly in line with the observational errors from the APOKASC2 and APOGEE datasets that the lower left panel is made from (see Appendix~\ref{app:obs_err} for plots of the observational errors). The error in \MgFe{} is not signficant, but the age error, which is $1\,\Gyr$ at $10\,\Gyr$, smears out the distribution before and after the dashed line. The $\alpha$-enhanced SoI still shows two separate populations in this plot, but they now significantly overlap in age.

@@

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{fig/392276_SFH_AGN_M200.pdf}
  \caption{\textbf{The evolutionary history of our subhalo of interest.} The upper panel shows the SFH of our subhalo. This SFH is generated using the formation times of all star particles in the subhalo at $z=0$. In this and all subsequent panels we show a vertical dashed line at the same position as in Figure~\ref{fig:alpha} ($\sim3.2\,\Gyr$), which delineates the separation between the high- and low-$\alpha$ sequences. The gap in ages is naturally associated with a gap in the SFH located at the dashed line. The middle panel shows the BH accretion rate as a fraction of the maximum (Eddington) accretion rate across time. The accretion rate is high early on, but then drops. The gap in ages (dashed line) is coincident with a local peak in the BH accretion rate. In the lower panel we show the mass assembly of the halo, given as the mass within the radius that encloses $200\times$ the mean density of the universe. In this plot mergers are shown as sudden increases in $M_{200}$. Mergers at very early times when the halo is signifcantly less massive than its $z=0$ mass are not shown, but one can see that no clear merger is associated with the gap in stellar ages. Two mergers occur later on at $t\sim10$ and $\sim12\,\Gyr$.}
  \label{fig:history}
\end{figure}

In an effort towards understanding the events in our SoI's history that led to the behavior around the dashed line in Figure~\ref{fig:alpha}, we examine the evolution of some of its key quantities. In the upper panel of Figure~\ref{fig:history}, we show the SFH of our galaxy with the same dashed line as in Figure~\ref{fig:alpha} (at an age of $10.6\,\Gyr$ in that plot is at a time of $\sim3.2\,\Gyr$ here). This SFH is computed from the formation times of all star particles in the subhalo at $z=0$. This allows us to compute the SFH at fine time resolution. We see two peaks in the SFH at $t\sim2.5$ and $\sim4.5\,\Gyr$ of an amplitude of about $70$ and $60\,\Msunyr$. The dashed line corresponds to a relative drop in the SFR, down to about $10\,\Msunyr$ before quickly recovering.

The middle panel shows the accretion rate of the central black hole as a fraction of the maximum (Eddington) accretion rate. Early in its history ($t<2\,\Gyr$), the subhalo experiences high accretion rates. The accretion rate then steadily declines until $t\sim5\,\Gyr$ at which point it switches to the low accretion radio mode. However, at the dashed line ($\sim3.2\,\Gyr$) there is a localized peak in the accretion rate, maxing out at about $30\%$ Eddington. This places the BH firmly in the thermal quasar-mode, and allows it to inject large amounts of thermal energy into the center of the galaxy.

@@

We show the evolution of the SoI in greater detail at the transition between the high- and low-$\alpha$ sequences in Figure~\ref{fig:history_zoom}. As in Figure~\ref{fig:history} we indicate the SFR (blue) and black hole accretion rate (orange), but we also include the strength of the bar, taken to be the maximum value of $A_2$, in red. From $\sim2.8\,\Gyr$ to $\sim3.2\,\Gyr$, the SFR drops from $\sim54$ to $3.6\,\Msunyr$. The minimum in the SFR is nearly coincident with the transition line.

@@

We also show four relevant statistics in each panel of Figures~\ref{fig:images_star} and \ref{fig:images_gas}. For Figure~\ref{fig:images_star}, these are, clockwise from the upper left, the SFR, cosmic time, bar strength, and BH accretion rate. They are the same for Figure~\ref{fig:images_gas}, but bar strength is replaced with gas fraction (within twice the stellar half-mass radius). The bar strength is missing from several snapshots. This is because the catalog we pull \Atmax{} from excludes certain galaxies from being considered barred if they do not pass certain quality cuts \citep{2022MNRAS.515.1524Z}. These can come from, e.g., flybys or violent buckling events.

@@

The age gap in each \FeH{} bin of Figure~\ref{fig:alpha} is contemperaneous with a minimum in the global SFR. We can see in the upper panel of Figure~\ref{fig:history} that this dashed line lies almost exactly at the point of a local minimum in the SFH. This minimum, which is $\sim10\,\Msunyr$, is 6--7$\times$ smaller than the maxima before and after it.\footnote{A reasonable argument can be made that $10\,\Msunyr$ is hardly quiescent. However, for our purposes, it only matters that the SFR at the minimum is small relative to the SFR close in time before and after.}

@@

\subsection{Steepening of $\alpha$ Decline}\label{ssec:sfe}
As described in Section~\ref{X}, we applied a post-processing to the \MgFe{} of star particles in the TNG simulation. Specifically, for star particles formed before $z=1.5$, we added to their \MgFe{} a value of $0.1\times\left(t_{1.5}-t_{\textrm{form}}\right)$, where $t_{1.5}$ is the age of the universe at $z=1.5$ ($\sim4.3\,\Gyr$) and $t_{\textrm{form}}$ is the formation time. This post-processed subhalo is presented alongside the fiducial subhalo in the right and middle columns, respectively, of Figures~\ref{fig:fig1} and \ref{fig:alpha}.

@@

A complicating factor for this picture comes from the high-SFR associated with the galaxy. The depletion time ($M_{\textrm{gas}}/\textrm{SFR}$) at the $t=2.84\,\Gyr$ snapshot is $204\,\Myr$, shorter than the time it takes for the SFR to reach its minimum. However, this perspective has to be balanced by the fact that the depletion time is already short at this redshift. The average depletion time in the preceding 5 snapshots is $220\,\Myr$, so clearly the SoI is accreting high amounts of gas from its CGM. A definitive account, difficult with the current simulation because of its sparse snapshot spacing, requires further work.

>>>>

\begin{abstract}
    The Milky Way is known to host at least two modes in its present day distribution of Fe and $\alpha$-elements. The exact cause of this bimodality is disputed, but one class of explanations involves the merger between the Milky Way and a relatively massive satellite (Gaia-Sausage-Enceladus) at $z\sim2$. However, reproducing this bimodality in simulations is not straightforward, with conflicting results on the prevalance, morphology, and mechanism behind multimodality. We present a case study of a galaxy in the Illustris TNG50 simulation which undergoes a sequence of starburst, brief quiescence, and then rejuvenation. After a minor post-processing step which boosts the \alphaFe{} of old star particles, we demonstrate that this galaxy hosts a strongly bimodal distribution in the \alphaFe{}-\FeH{} plane. The high- and low-$\alpha$ sequences are neatly separated in time by the brief quiescent period. The quiescent period in this galaxy is not associated with a merger but by the formation of a bar followed by AGN activity. This galaxy indicates a novel scenario in which the $\alpha$-bimodality is caused by AGN-induced quenching triggered by the formation of the Milky Way's bar. We argue that the post-processing step can be interpreted as the TNG model underproducing star formation in the densest regions at high-$z$.
  \end{abstract}

@@

In that work, we used idealized simulations of a galaxy merger that triggered a starburst which preceded the quiescent period. However, we argued that the merger aspect of that work was not necessary, but rather the quiescence. In this work, we study a subhalo from the Illustris TNG50 simulation which demonstrates this. This subhalo of interest (SoI) exhibits the sequence of events presented in \citet{2024arXiv240707985B} after a simple post-processing step which enhances the \alphaFe{} of old star particles. The SoI undergoes a brief quiescent period which neatly separates a high- and low-$\alpha$ sequence. However, instead of being preceded by a merger, the quenching is preceded by apparent bar-induced AGN activity. Therefore, this work serves as a verification that the scenario in \citet{2024arXiv240707985B} is possible in cosmological simulations and can occur in non-merger scenarios.

In Section~\ref{sec:methods} we discuss our selection technique which led to discovering the SoI, the observations we use for comparison, as well as a simple one zone chemical evolution model we use to justify our post-processing step. In Section~\ref{sec:results} we present the main results which we discuss and interpret in Section~\ref{sec:disc}. We conclude in Section~\ref{sec:conc}.

@@

Using the public catalog, we selected a sample of subhalos at $z=1.5$ (snapshot 40) according to the following criteria: (1) the subhalo is central (i.e., the most massive subhalo within its halo), and (2) the subhalo's stellar mass is between $10^{10}$ and $10^{10.5}\,\Msun/\textrm{h}$. There were a total of 168 subhalos that met both criteria. The chosen mass range is broadly consistent with the expected mass of the Milky Way at this redshift \citep{2013ApJ...771L..35V}. We chose to make our selection of galaxies at $z=1.5$ instead of at lower redshift because we wished to capture the \textit{formation} of any multimodal structure. We did not want contamination by mergers at lower redshift which we know contribute very little to the Milky Way's disk stars \citep[e.g.,][]{2016ARA&A..54..529B}.

@@

\subsection{Bar Catalog}\label{ssec:barcat}
In order to study the evolution of the morphology of our SoI, we make use of the \citet{2022MNRAS.515.1524Z} catalog. They provide a decomposition of all galaxies with at least $10^4$ star particles, as well as an estimate of the bar strength. This taken to be the maximum of the second Fourier mode ($A_{2,\textrm{max}}$) within the bar region (not including spiral arms).

@@

The colored vertical lines are indicated at $\FeH=-0.75$, $-0.5$, $-0.25$, and $0\,\dex$ in the Milky Way, and at bins $0.25\,\dex$ higher in the simulations. The lower panels show 1D histograms of \MgFe{} in bins centered on these values. The bins for the Milky Way/simulations have width $0.2$/$0.05\,\dex$. The \MgFe{} values are given offsets so their medians are at zero. The rationale for the higher plotted \FeH{} in the simulations comes just from the empirical location of the bimodalities. The Milky Way shows a clear bimodal population, with a high-$\alpha$ sequence most clearly distinct from the low-$\alpha$ sequence at low metallicity. The two sequences merge around solar metallicity.

@@

In the lower panels we plot to the left data from the Milky Way and to the middle and right from the fiducial and $\alpha$-enhanced SoI, respectively. For the simulations, we add $15\%$ age and $0.01\,\dex$ \MgFe{} Gaussian errors. These values are roughly in line with the observational errors from the APOKASC2 and APOGEE datasets that the lower left panel is made from (see Appendix~\ref{app:obs_err} for plots of the observational errors). The error in \MgFe{} is not signficant, but the age error, which is $1.5\,\Gyr$ at $10\,\Gyr$, smears out the distribution before and after the dashed line. The $\alpha$-enhanced SoI still shows two separate populations in this plot, but they now significantly overlap in age.

@@

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{fig/392276_SFH_AGN_M200.pdf}
  \caption{\textbf{The evolutionary history of our subhalo of interest.} The upper panel shows the SFH of our subhalo. This SFH is computed using the SFR within twice the stellar half-mass radius in our SoI's main progenitors. In this and all subsequent panels we show a vertical dashed line at the same position as in Figure~\ref{fig:alpha} ($\sim3.2\,\Gyr$), which delineates the separation between the high- and low-$\alpha$ sequences. The gap in ages is naturally associated with a gap in the SFH located at the dashed line. The middle panel shows the BH accretion rate as a fraction of the maximum (Eddington) accretion rate across time. The accretion rate is high early on, but then drops. The gap in ages (dashed line) is coincident with a local peak in the BH accretion rate. In the lower panel we show the mass assembly of the halo, given as the mass within the radius that encloses $200\times$ the mean density of the universe. In this plot mergers are shown as sudden increases in $M_{200}$. Mergers at very early times when the halo is signifcantly less massive than its $z=0$ mass are not shown, but one can see that no clear merger is associated with the gap in stellar ages. Two mergers occur later on at $t\sim10$ and $\sim12\,\Gyr$.}
  \label{fig:history}
\end{figure}

In an effort towards understanding the events in our SoI's history that led to the behavior around the dashed line in Figure~\ref{fig:alpha}, we examine the evolution of some of its key quantities. In the upper panel of Figure~\ref{fig:history}, we show the SFH of our galaxy with the same dashed line as in Figure~\ref{fig:alpha} (at an age of $10.6\,\Gyr$ in that plot is at a time of $\sim3.2\,\Gyr$ here). This SFH is computed using the SFR within twice the stellar half-mass radius in our SoI's main progenitors. We see two peaks in the SFH at $t\sim2.5$ and $\sim4.5\,\Gyr$ of an amplitude of about $50$ and $30\,\Msunyr$. The dashed line corresponds to a relative drop in the SFR, down to about $3\,\Msunyr$ before quickly recovering.

The middle panel shows the accretion rate of the central black hole as a fraction of the maximum (Eddington) accretion rate. Early in its history ($t<2\,\Gyr$), the subhalo experiences high accretion rates. The accretion rate then steadily declines until $t\sim5\,\Gyr$ at which point it fails to show on this linear plot. However, at the dashed line ($\sim3.2\,\Gyr$) there is a localized peak in the accretion rate, maxing out at about $30\%$ Eddington. This places the BH firmly in the thermal quasar-mode, and allows it to inject large amounts of thermal energy into the center of the galaxy.

@@

We show the evolution of the SoI in greater detail at the transition between the high- and low-$\alpha$ sequences in Figure~\ref{fig:history_zoom}. As in Figure~\ref{fig:history}, we indicate the SFR (blue) and black hole accretion rate (orange), but we also include the strength of the bar, taken to be the maximum value of $A_2$ in the bar region, in red. From $\sim2.8\,\Gyr$ to $\sim3.2\,\Gyr$, the SFR drops from $\sim54$ to $3.6\,\Msunyr$. The minimum in the SFR is nearly coincident with the transition line.

@@

We also show four relevant statistics in each panel of Figures~\ref{fig:images_star} and \ref{fig:images_gas}. For Figure~\ref{fig:images_star}, these are, clockwise from the upper left, the SFR, cosmic time, bar strength, and BH accretion rate. They are the same for Figure~\ref{fig:images_gas}, but bar strength is replaced with gas fraction (within twice the stellar half-mass radius). The bar strength is missing from several snapshots because the \citet{2022MNRAS.515.1524Z} catalog excludes certain galaxies from being considered barred if they do not pass certain quality cuts. These can come from, e.g., flybys or violent buckling events.

@@

The age gap in each \FeH{} bin of Figure~\ref{fig:alpha} is contemperaneous with a minimum in the global SFR. We can see in the upper panel of Figure~\ref{fig:history} that this dashed line lies almost exactly at the point of a local minimum in the SFH. This minimum, which is $\sim3\,\Msunyr$, is 10--15$\times$ smaller than the maxima before and after it.

@@

\subsection{Steepening of $\alpha$ Decline}\label{ssec:sfe}
As described in Section~\ref{ssec:tng}, we applied a post-processing to the \MgFe{} of star particles in the TNG simulation. Specifically, for star particles formed before $z=1.5$, we added to their \MgFe{} a value of $0.1\times\left(t_{1.5}-t_{\textrm{form}}\right)$, where $t_{1.5}$ is the age of the universe at $z=1.5$ ($\sim4.3\,\Gyr$) and $t_{\textrm{form}}$ is the formation time. This post-processed subhalo is presented alongside the fiducial subhalo in the right and middle columns, respectively, of Figures~\ref{fig:fig1} and \ref{fig:alpha}.

@@

A complicating factor for this picture comes from the high SFR associated with the galaxy. The depletion time ($M_{\textrm{gas}}/\textrm{SFR}$) at the $t=2.84\,\Gyr$ snapshot is $204\,\Myr$, shorter than the time it takes for the SFR to reach its minimum. However, this has to be balanced by the fact that the depletion time is already short at this redshift. The average depletion time in the preceding 5 snapshots is $220\,\Myr$, so clearly the SoI is accreting high amounts of gas from its CGM. A definitive account, difficult with the current simulation because of its sparse snapshot spacing, requires further work.
